---
title: "Vignette - Zip Code Coordinate Data Preparation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette - Zip Code Coordinate Data Preparation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  purl = FALSE
)
```

```{r, message = FALSE, warning = FALSE , eval = FALSE, purl = FALSE}
library(tidyverse)
library(disperseR)
```

In this vignette we show you how we prepared the zip code coordinate data that we have attached to the `disperseR` package. 

In case you have not yet created your project folder use the following command. 

```{r, eval = FALSE, purl = FALSE}
disperseR::create_dirs()
```

We first download the data. The original source was Open Data Soft's
US ZIP code latitude/longitude dataset. The data is now bundled in the package
as `disperseR::zipcodecoordinate`, so manual download is optional.

```{r, eval = FALSE, purl = FALSE}
filename <- file.path(main_dir, 'zipcodecoordinate.csv')
url <- 'https://public.opendatasoft.com/api/explore/v2.1/catalog/datasets/us-zip-code-latitude-and-longitude/exports/csv'

if(!file.exists(filename)){
    download.file(url = url, destfile = filename)
}
```


Read in the file. 

```{r, eval = FALSE, purl = FALSE}
zipcodecoordinate <- read.csv(filename, skip = 0, header = T, sep=";")
```


Select columns and rename them. 


```{r, eval = FALSE, purl = FALSE}
zipcodecoordinate <- zipcodecoordinate %>% 
  dplyr::select(Zip, City, Latitude, Longitude) %>% 
  dplyr::rename(ZIP=Zip)
```

Now the data set is ready to use in your plots! We added it to this package. 
